---
title: <center>Pedigree Drawing</center>
author: <center>Bruce Cochrane</center>
date: <center>`r format(Sys.Date(),format="%B %d %Y")`</center>
output: 
    html_document:
       css: Chapter.css
       toc: yes
       toc_depth: 5
---

 the package kinship2
```{r}
#install.packages("kinship2")
```
```{r}
library(kinship2)
```
First, use data supplied to get format
```{r}
data(minnbreast)
head(minnbreast)
```

So I wonder if we could create a dataframe via excel that just has the first four columns (actually five, including sex)

Note that when R 3.2.2 is released, we should be able to read https and thus get to Dropbox directly.  For now, we'll stick with direct file loading.
```{r}
#dat.raw <-read.csv("https://dl.dropboxusercontent.com/u/9752688/QPopgen/Summer15Development/DATA/SimplePedigree.csv",stringsAsFactors=FALSE)
dat.raw <-read.csv("./scratch/SimplePedigree.csv",stringsAsFactors=FALSE)
head(dat.raw)
```
Now create a pedigree object
```{r}
dat.ped <-with(dat.raw,pedigree(id,fatherid,motherid,sex))
plot(dat.ped)
```
That is super cool!!  We can use it in genetics next fall for all types of stuff.  And for just drawing a structure, all we need is three columns - motherid, fatherid and sex.  We're going to make a file for that
```{r}
dat.raw <-dat.raw[,c(1,3,4,6)]
dat.raw
```
```{r}
save(dat.raw,file="../Data2015/Pedigree")
```
### May 8

OK.  now we want to add in affected status.  We'll just do it randomly for now
```{r}
aff <-rbinom(12,1,.5)
aff
```
And we'll cbind that to the data
```{r}
dat.new <-cbind(dat.raw,aff)
head(dat.new)
```
And now we should be able to generate the file
```{r}
ped.2 <-with(dat.new, pedigree(id,fatherid,motherid,sex,aff))
plot(ped.2)
```
Well what do you know?  That actually works as an autosomal dominant!!  Of course, we didn't set.seed, so we'll never get it again . . . 


Try it with names
```{r}
dat.me <-read.csv("./scratch/MyFamily.csv")
dat.me
```
```{r}
ped.me <-with(dat.me,pedigree(id,fatherid,motherid,sex,missid="0"))
plot(ped.me)
```
### For class.

So what we need to do is to develop csv files for the ususal suspects - dominant, recessive, x linked.  Also would like to have ones for ambiguous cases and, if we can figure it out, a lod score example.  And I wonder if using names is better than numbers.

### Waardenberg

```{r}
dat.ws <-read.csv("./scratch/WS3.csv")
head(dat.ws)
```
```{r}
ped.ws <-with(dat.ws,pedigree(id,dadid=fatherid,momid=motherid,sex=sex,affected=aff,missid="0",))
plot(ped.ws,cex=.8)
```
Now, I think we could add a relationship matrix - just out of curiosity
```{r}
rel <-data.frame(id1="4-14",id2="4-15",code=2)
rel
```
OK.  This is going to be hokey (since it uses attach) but we'll see if it works.

```{r}
#attach(dat.ws)
ped2.ws <-pedigree(id,dadid=fatherid,momid=motherid,sex=sex,affected=aff,missid="0",relation = rel)
plot(ped2.ws,cex=.8)
```
OK.  It works. But we need to be better than this.  First, let's see if we can generate random names:

```{r}
install.packages("randomNames")
```
```{r}
#library(randomNames)
test <-randomNames(40,which.names="first",gender=1,ethnicity=5
)
test
```
OK.  So we can get a list of males with gender=0 and females with gender =1.

Now, we've done that for an autosomal dominant
```{r}
dat.rf <-read.csv("./scratch/RoyalFamily.csv")
dat.rf
```
Now the pedigree.
```{r}
ped.rf <-with(dat.rf,pedigree(id,dadid=fatherid,momid=motherid,sex=sex,affected=aff,missid="0"))
plot(ped.rf,cex=.60)
```
So all in all, not all that bad.  One more for the road - sex-linked recessive.  And make them 

```{r}
dat.xr <-read.csv("./scratch/XRec.csv")
ped.xr <-with(dat.xr,pedigree(id,dadid=fatherid,momid=motherid,sex=sex,affected=aff,missid="0"))
plot(ped.xr,cex=.6)
```
Just for future reference, we can extract data from the package with commmands like:
```{r}
nms.male.na <-randomNamesData$first_names_e3_g1
```
```{r}
str(nms.male.na)
```
```{r}
names(nms.male.na)[which(nms.male.na==max(nms.male.na))]
```
```{r}
ls.str(randomNamesData)
```

